---
title: "07_ModelSelection2"
---

```{r}
#| label: load_packages
#| echo: false

# Load required packages

pacman::p_load(
  ## data manipulation
  dplyr, tibble, tidyverse, broom, broom.mixed,
  
  ## model fitting
  ape, arm, brms, broom.mixed, cmdstanr, emmeans, glmmTMB, MASS, phytools, rstan, TreeTools,
  
  ## model checking and evaluation
  DHARMa, loo, MuMIn, parallel,
  
  ## visualisation
  bayesplot, ggplot2, patchwork, tidybayes,
  
  ## reporting and utilities
  gt, here, kableExtra, knitr
)
```

## Example 2

We revisit the dataset on Estrildid finches to examine how food supplementation affects gaze frequency toward dot patterns (as introduced in `Beyond Gaussian 1` section).

To avoid repeating the structure of the previous example, we now take a more realistic modeling approach by incorporating phylogenetic non-independence as a random effect. The response variable, frequency, is a count variable, which supports the use of Poisson or Negative Binomial distributions. However, as visualised in the plots below, there is substantial between-species variation—both in mean gaze frequency and in the degree of within-species variability.

To better capture this structure, we include phylogeny as a random effect, distinct from the previously used species effect. While species was treated as a simple categorical grouping factor, the phylogenetic random effect accounts for shared evolutionary history and allows for correlated responses among closely related species.

In the following sections, we will:

-   Include phylogeny as a random effect in the location and/or scale parts of the model.

-   Compare models with and without phylogenetic structure to evaluate whether accounting for evolutionary history improves model fit.

::: panel-tabset

### Step 1

#### Identify data type and plot raw data

The violin plot by condition and sex shows that gaze frequency is generally higher in the food-deprived condition, with some individuals—especially females—showing very high values (over 500). The faceted line plot by species shows: 

-   Consistent directional trends (e.g., reduced gazing in the supplied condition)
-   Species-specific differences in variance, suggesting possible heteroscedasticity

These patterns raise two important considerations: 

1. Overdispersion is likely present, particularly in the deprived condition, where the spread of counts is large. 
2. A phylogenetic effect may be at play, since species vary not only in their average gaze frequencies but also in how consistent individuals are within each species - possibly due to traits shaped by common ancestry.

```{r}
#| label: identify_data_type - eg2
#| fig-width: 8
#| fig-height: 6

ggplot(dat_pref, aes(x = condition,
                     y = frequency)) +
    geom_violin(color = "#8B8B83", fill = "white",
              width = 1.2, alpha = 0.3) +
  geom_jitter(aes(shape = sex, color = sex),
              size = 3, alpha = 0.8,
              width = 0.15, height = 0) +
  labs(
    title = "Total frequency of gazes towards dot patterns",
    x = "Condition",
    y = "Total frequency of gazes (1hr)"
  ) +
  scale_shape_manual(values = c("M" = 17, "F" = 16),
                     labels = c("M" = "Male", "F" = "Female")) +
  scale_color_manual(values = c("M" = "#009ACD", "F" = "#FF4D4D"),
                     labels = c("M" = "Male", "F" = "Female")) +
  theme_classic(base_size = 16) +
  theme(
    axis.text = element_text(color = "#6E7B8B", size = 14),
    axis.title = element_text(color = "#6E7B8B", size = 14),
    legend.title = element_text(color = "#6E7B8B"),
    legend.text = element_text(color = "#6E7B8B"),
    legend.position = "right",
  axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(dat_pref, aes(x = condition, y = frequency)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Total frequency of gazes towards dot patterns by species",
    x = "Condition",
    y = "Total frequency of gazes (1hr)"
  ) +
  stat_summary(fun = mean, geom = "line", aes(group = species, color = species)) +
  theme_classic() +
  facet_wrap(~ species)
```

### Steps 2 and 3

#### Begin with a simpler model and check residual diagnostics

From the previous analysis, we already know that a location-only model without phylogenetic effects does not adequately capture the structure of the data. However, here we re-examine this model before moving on.

```{r}
#| label: model_fitting - eg2

model_0 <- glmmTMB(
  frequency ~ 1 +  condition + sex + (1 | species) + (1 | id), # location part (mean)
  data = dat_pref,
  family = nbinom2(link = "log")
)

```

To evaluate model adequacy, we use the `DHARMa` package to simulate and visualise residuals. The plots allow us to assess residual uniformity, potential over- or underdispersion, outliers, and leverage.

```{r}
#| label: residual_diagnostics - eg2
# main diagnostic plots

model0_res <- simulateResiduals(model_0, plot = TRUE, seed = 42)

# formal test for over/underdispersion
testDispersion(model0_res) 
```

Although the dispersion test (`p = 0.456`) suggests no clear evidence of overdispersion based on the mean–variance relationship, the KS test indicates that the residuals deviate from a uniform distribution. These two tests assess different aspects of model adequacy: the dispersion test evaluates whether the variance is correctly specified as a function of the mean, while the KS test can detect broader misfits, such as unmodeled structure, zero inflation, or group-specific heteroscedasticity. Thus, even in the absence of overdispersion, the presence of non-uniform residuals supports the need for a more flexible model.

Since we have found the limitations of the model without phylogenetic effects, we now proceed to fit the same model using the `brms` package, this time including the phylogenetic effect as a random effect to account for the non-independence among species. Including both species and phylogeny as random effects allows us to distinguish between two sources of variation: the phylogenetic effect captures correlations arising from shared evolutionary history, whereas the species-level random intercept absorbs residual species-specific variation not accounted for by the tree—such as ecological or methodological factors. Including both terms helps avoid underestimating heterogeneity at the species level and provides a more accurate partitioning of variance.

First, we run the same model as above, but using the `brms` package. Then, we will gradually increase the model complexity by adding phylogenetic effects and a scale part to the model.

```{r}
#| label: model_fitting1 - eg2
#| eval: false

formula_eg2.0 <- bf(
  frequency ~ 1 + condition + sex + (1 | species) + (1 | id)
)

prior_eg2.0 <- default_prior(formula_eg2.0, 
                        data = dat_pref, 
                        family = negbinomial(link = "log")
)

model_eg2.0 <- brm(formula_eg2.0, 
            data = dat_pref, 
            prior = prior_eg2.0,
            chains = 2, 
            iter = 5000, 
            warmup = 3000,
            thin = 1,
            family = negbinomial(link = "log"),
            save_pars = save_pars(all = TRUE)
            # control = list(adapt_delta = 0.95)
)

```

Then, we add the phylogenetic effect as a random effect to the model. This allows us to account for the non-independence among species due to shared evolutionary history.

```{r}
#| label: model_fitting2 - eg2
#| eval: false

tree <- read.nexus(here("data", "AM_est_tree.txt"))
tree　<-force.ultrametric(tree, method = "extend")
# is.ultrametric(tree)

tip <- c("Lonchura_cantans","Uraeginthus_bengalus","Neochmia_modesta",
         "Lonchura_atricapilla","Padda_oryzivora","Taeniopygia_bichenovii",
         "Emblema_pictum","Neochmia_ruficauda","Lonchura_maja",
         "Taeniopygia_guttata","Chloebia_gouldiae","Lonchura_castaneothorax")
tree <- KeepTip(tree, tip, preorder = TRUE, check = TRUE)

# Create phylogenetic correlation matrix
A <- ape::vcv.phylo(tree, corr = TRUE)

# Specify the model formula (location-only)
formula_eg2.1 <- bf(
  frequency ~ 1 + condition + sex + 
    (1 | a | gr(phylo, cov = A)) +  # phylogenetic random effect
    (1 | species) +             # non-phylogenetic species-level random effect (ecological factors)
    (1 | id)                    # individual-level random effect
)

prior_eg2.1 <- brms::get_prior(
  formula = formula_eg2.1, 
  data = dat_pref, 
  data2 = list(A = A),
  family = negbinomial(link = "log")
)

model_eg2.1 <- brm(
  formula = formula_eg2.1, 
  data = dat_pref, 
  data2 = list(A = A),
  chains = 2, 
  iter = 12000, 
  warmup = 10000,
  thin = 1,
  family = negbinomial(link = "log"),
  prior = prior_eg2.1,
  control = list(adapt_delta = 0.95),
  save_pars = save_pars(all = TRUE)
)
```

### Step 4

#### Gradually increase model complexity

In the previous model, variation in individual gaze frequency was modeled assuming a constant dispersion parameter across all groups. However, visual inspection of the raw data suggests heteroscedasticity (i.e., group-specific variance). To accommodate this, we now fit a location-scale model where the shape (dispersion) parameter is allowed to vary by condition and sex.

This model helps capture structured variability in the data and may reveal whether the precision (or variability) of gaze frequency differs systematically between experimental groups.

```{r}
#| eval: false

# Specify the model formula (location-scale) - the scale part does not include random effects
formula_eg2.2 <- bf(
  frequency ~ 1 + condition + sex + 
    (1 | a | gr(phylo, cov = A)) +  
    (1 | species) +          
    (1 | id),              
shape ~ 1 + condition + sex
)

prior_eg2.2 <- brms::get_prior(
  formula = formula_eg2.2, 
  data = dat_pref, 
  data2 = list(A = A),
  family = negbinomial(link = "log", link_shape = "log")
)

model_eg2.2 <- brm(
  formula = formula_eg2.2, 
  data = dat_pref, 
  data2 = list(A = A),
  chains = 2, 
  iter = 12000, 
  warmup = 10000,
  thin = 1,
  family = negbinomial(link = "log", link_shape = "log"),
  prior = prior_eg2.2,
  control = list(adapt_delta = 0.95),
  save_pars = save_pars(all = TRUE)
)

```

```{r}
#| eval: false

# specify the model formula (location-scale)
## the scale part includes phylogenetic random effects and individual-level random effects

formula_eg2.3 <- bf(
  frequency ~ 1 + condition + sex + 
    (1 |a| gr(phylo, cov = A)) +  # phylogenetic random effect
    (1 | species) +            # non-phylogenetic species-level random effect (ecological factors)
    (1 | id),              # individual-level random effect
shape ~ 1 + condition + sex +
    (1 |a| gr(phylo, cov = A)) +  
    (1 | species) +  
    (1 | id)          
    )

prior_eg2.3 <- brms::get_prior(
  formula = formula_eg2.3, 
  data = dat_pref, 
  data2 = list(A = A),
  family = negbinomial(link = "log", link_shape = "log")
)

model_eg2.3 <- brm(
  formula = formula_eg2.3, 
  data = dat_pref, 
  data2 = list(A = A),
  chains = 2, 
  iter = 12000, 
  warmup = 10000,
  thin = 1,
  family = negbinomial(link = "log", link_shape = "log"),
  prior = prior_eg2.3,
  control = list(adapt_delta = 0.95),
  save_pars = save_pars(all = TRUE)
)

summary(model_eg2.3)

```

### Step 5

#### Compare models using information criteria

To assess whether model fit improves with increased complexity, we compare models using Leave-One-Out Cross-Validation (LOO).

```{r}
#| eval: false

options(future.globals.maxSize = 2 * 1024^3)
loo_eg2.0 <- loo::loo(model_eg2.0, moment_match = TRUE,
                      reloo = TRUE, cores = 2)
loo_eg2.1 <- loo::loo(model_eg2.1, moment_match = TRUE, 
                      reloo = TRUE, cores = 2)
loo_eg2.2 <- loo::loo(model_eg2.2, moment_match = TRUE,
                      reloo = TRUE, cores = 2)
loo_eg2.3 <- loo::loo(model_eg2.3, moment_match = TRUE,
                      reloo = TRUE, cores = 2)

fc_eg2 <- loo::loo_compare(loo_eg2.0, loo_eg2.1, loo_eg2.2, loo_eg2.3)

print(fc_eg2)
#             elpd_diff se_diff
# model_eg2.3   0.0       0.0  
# model_eg2.1 -11.5       4.5  
# model_eg2.2 -11.7       5.3  
# model_eg2.0 -13.5       4.9  

summary(model_eg2.3)
```

Model eg2.3, which includes both phylogenetic and species-level effects on the location and scale parts, performs best (highest elpd).

ELPD stands for Expected Log Predictive Density. It is a measure of out-of-sample predictive accuracy—that is, how well a model is expected to predict new, unseen data. 
-  It is used in Bayesian model comparison, especially when using LOO-CV (Leave-One-Out Cross-Validation). 
-  Higher ELPD means better predictive performance. 
-  It is calculated by taking the log-likelihood of each observation, averaged over posterior draws, and then summed over all observations. 
-  It rewards models that fit the data well without overfitting.

```{r}
#| echo: false

model_eg2.3 <- readRDS(here("Rdata", "model_eg2.3.rds"))
summary(model_eg2.3)

```

Accounting for phylogeny captures much of the between-species structure, it highlights evolutionary constraints or shared ecological traits. The remaining species-specific variance suggests additional factors (e.g., experimental nuances) not captured by the tree alone.
:::
